---
title: "Clustering"
output: html_document
date: "2025-10-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load libraries
install.packages("patchwork")
install.packages("plotly")
library(tidyverse)
library(cluster)
library(ggplot2)
library(patchwork)
library(plotly)
```

## Task 1 -- Clusters and the Gap Statistic
Here is a function to generate the data:
```{r}
generate_hypercube_clusters <- function(n, k, side_length, noise_sd = 1.0) {
  points_list <- list()
  labels <- c()
  
  # creating a matrix (n x n) with side length on the diagonal, zeroes otherwise
  # each row is a cluster center at a positive corner of the cube
  # the positive corners are exactly the points where one coordinate = side_length, all others = 0
  cluster_centers <- diag(side_length, n, n)
  
  for (i in 1:n) {
    # Generate k points around this center with Gaussian noise
    cluster_points <- matrix(rnorm(k * n, mean = 0, sd = noise_sd), nrow = k, ncol = n) + 
                      matrix(rep(cluster_centers[i, ], each = k), nrow = k)
    points_list[[i]] <- cluster_points
    labels <- c(labels, rep(i, k))
  }
  
  points <- do.call(rbind, points_list)
  return(list(points = points, labels = labels))
}
```

Now, let's use the function above to create datasets where clusters move progressively closer. Running a simulation below.
```{r}
# parameters for the simulation below
k <- 100 # points per cluster
noise_sd <- 1.0 # keep noise at 1 sd
dimensions <- c(6,5,4,3,2) # n
side_lengths <- 10:1 #for each dimension, we need to test side lengths from 10 down to 1

# for storing results
results <- data.frame()

# Loop over dimensions and side lengths
for (n in dimensions) {
  for (L in side_lengths) {
    cat("n =", n, "side_length =", L, "\n")
    
    # Generate data using my function above
    data <- generate_hypercube_clusters(n = n, k = k, side_length = L, noise_sd = noise_sd)
    points <- data$points
    
    # calculating the gap statistic with clusGap! using kmeans as our input function
    # specify nstart = 20 and iter.max = 50 to ensure good clustering
    gap_stat <- clusGap(points, FUN = function(x, k) {
                    kmeans(x, centers = k, nstart = 20, iter.max = 50)
                  },
                  K.max = n+2, B = 20)
    
    # let's find the best number of clusters according to Gap statistic
    optimal_k <- maxSE(gap_stat$Tab[,"gap"], gap_stat$Tab[,"SE.sim"], method = "firstSEmax")
    
    # Store results here
    results <- rbind(results, data.frame(n = n, side_length = L, estimated_k = optimal_k))
  }
}

# View results
print(results)
```

Interesting, let's do some visualization to understand what is happening here. Creating plots showing the estimated number of clusters as a function of the side length for each dimension.
```{r}
# make n as integer for plotting
results$n <- as.integer(results$n)

# List to store plots
plot_list <- list()

for (dim in sort(unique(results$n), decreasing = TRUE)) {
  df <- subset(results, n == dim)
  
  # Separate dataframe for the red reference line
  ref_line <- data.frame(yintercept = dim, label = "True number of clusters")
  
  p <- ggplot(df, aes(x = side_length, y = estimated_k, group = 1)) +
    geom_line(color = "steelblue", linewidth = 1.2) +
    geom_point(color = "steelblue", size = 2) +
    # Now use data=ref_line for the red reference line
    geom_hline(data = ref_line, aes(yintercept = yintercept, color = label),
               linetype = "dashed", size = 1) +
    scale_color_manual(values = c("True number of clusters" = "red")) +
    scale_x_reverse(breaks = 1:10) +
    coord_cartesian(ylim = c(0, 8)) +
    labs(
      title = paste("Dimensions (n):", dim),
      x = "side length",
      y = "Estimated # of clusters",
      color = ""  # Legend title
    ) +
    theme_minimal() +
    theme(
      text = element_text(size = 10),
      strip.text = element_text(size = 10),
      legend.position = "bottom"
    )
  
  plot_list[[as.character(dim)]] <- p
}

# Combine into a grid: 3 in top row, 2 in bottom
combined_plot <- (plot_list[["6"]] | plot_list[["5"]] | plot_list[["4"]]) / 
                 (plot_list[["3"]] | plot_list[["2"]]) +
                 plot_layout(guides = "collect")

combined_plot

#ggsave("figures/combined_plot.png", plot = combined_plot, width = 10, height = 6, dpi = 300)


```

Results for each dimension:
For n = 6, the estimated number of clusters is reduced at side length = 4.
For n = 5, the estimated number of clusters is reduced at side length = 3.
For n = 4, the estimated number of clusters is reduced at side length = 2.
For n = 3, the estimated number of clusters is reduced at side length = 2.
For n = 2, the estimated number of clusters is reduced at side length = 1.

My takeaway is that as the number of dimensions increase, it is crucial to have larger separation between cluster centers (larger side length values) to reliably detect the true number of clusters.


## Task 2 -- Spectral Clustering on Concentric Shells

First, let's build a function to generate shell clusters!
```{r}
generate_shell_clusters <- function(n_shells, k_per_shell, max_radius, noise_sd = 0.1) {
  # radius values from 1 to max evenly spaced for num of shells needed
  radius_values <- seq(1, max_radius, length.out = n_shells)
  
  # for storing results
  all_points <- data.frame()
  
  for (r in radius_values) {
    # let's get random points on sphere 
    theta <- runif(k_per_shell, 0, 2 * pi)
    phi <- acos(runif(k_per_shell, -1, 1))
    
    # add noise to radius, thickness to each shell
    noisy_r <- r + rnorm(k_per_shell, mean = 0, sd = noise_sd)
    
    # get 3D position for each point on the shell for plotting purposes
    x <- noisy_r * sin(phi) * cos(theta)
    y <- noisy_r * sin(phi) * sin(theta)
    z <- noisy_r * cos(phi)
    
    # make small df for one shell and then combine with previous shells
    shell_df <- data.frame(x, y, z, shell = as.factor(r))
    all_points <- rbind(all_points, shell_df)
  }
  
  return(all_points)
}
```

Now, let's generate a sample dataset and create an interactive 3D scatter plot using plotly to confirm this structure.
```{r}
set.seed(42)

# randomly picking param values for sample dataset
sample_data <- generate_shell_clusters(n_shells = 3, k_per_shell = 400, max_radius = 3)

# relabel shells for legend in plot
sample_data$shell_label <- factor(paste("Shell", 1:nlevels(sample_data$shell))[as.integer(sample_data$shell)])

# 3D scatter plot
plot_ly(sample_data, 
        x = ~x, y = ~y, z = ~z, 
        color = ~shell_label, 
        colors = "Set2",
        type = "scatter3d",
        mode = "markers",
        marker = list(size = 3)) %>%
  layout(
    title = list(text = "3D Concentric Shells test data", font = list(size = 16)),
    scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Z")
    ),
    legend = list(title = list(text = "Shell"))
  )
```


Great, now we can tell that our function works! Now we can implement spectral clustering. Inputs should be data (x) and number of clusters (k).
```{r}
spectral_clustering <- function(x, k, d_threshold = 1) {
  # Step 1: Build adjacency matrix
  dist_mat <- as.matrix(dist(x))
  A <- (dist_mat < d_threshold) * 1
  diag(A) <- 0

  # Step 2: Compute degree and Laplacian matrices
  D <- diag(rowSums(A))
  L <- D - A

  # Step 3: Compute normalized Laplacian L_sym = D^{-1/2} * L * D^{-1/2}
  D_inv_sqrt <- diag(1 / sqrt(diag(D)))
  D_inv_sqrt[!is.finite(D_inv_sqrt)] <- 0  # handles isolated points safely
  L_sym <- D_inv_sqrt %*% L %*% D_inv_sqrt

  # Step 4: Eigen decomposition
  eig <- eigen(L_sym, symmetric = TRUE)

  # smallest k non-zero eigenvectors
  tol <- 1e-10
  nonzero_idx <- which(eig$values > tol)       # filter out numerical zeros
  eig_order <- order(eig$values[nonzero_idx])  # ascending order of non-zero eigenvalues
  U <- eig$vectors[, nonzero_idx[eig_order[1:k]]]  # pick k smallest non-zero eigenvectors

  # K-means on rows of U
  # normalize U?
  U <- U / sqrt(rowSums(U^2))
  U[!is.finite(U)] <- 0
  km <- kmeans(U, centers = k, nstart = 20, iter.max = 200)

  return(km)
}


```

Great, now we have our function generate_shell_clusters() and can use this to run a simulation.
```{r}
# parameters for simulation
radius_values <- seq(10, 0, by = -1)
# constants below
n_shells <- 4
k_per_shell <- 100
noise_sd <- 0.1

results_shells <- data.frame(max_radius = numeric(), estimated_k = numeric())

for (r in radius_values) {
  # generate dataset
  data_shells <- generate_shell_clusters(
    n_shells = n_shells,
    k_per_shell = k_per_shell,
    max_radius = r,
    noise_sd = noise_sd
  )
  
  # just extract the points
  points <- as.matrix(data_shells[, c("x", "y", "z")])
  
  #set.seed(123)
  # run gap statistic using spectral clustering
  gap <- clusGap(points,
               FUN = function(x, k){ spectral_clustering(x, k, d_threshold = 1)},
               K.max = n_shells + 2,
               B = 20)

  
  est_k <- maxSE(gap$Tab[,"gap"], gap$Tab[,"SE.sim"], method = "firstSEmax")
  
  results_shells <- rbind(results_shells, data.frame(max_radius = r, estimated_k = est_k))
}

results_shells
```

Now, let's plot the estimated number of clusters vs. maximum radius!
```{r}
# Make sure max_radius is numeric for plotting
results_shells$max_radius <- as.numeric(results_shells$max_radius)

# Reference line for true number of clusters
true_k <- 4
ref_line <- data.frame(yintercept = true_k, label = "True number of clusters")

# Plot
ggplot(results_shells, aes(x = max_radius, y = estimated_k, group = 1)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(data = ref_line, aes(yintercept = yintercept, color = label),
             linetype = "dashed", size = 1) +
  scale_color_manual(values = c("True number of clusters" = "red")) +
  scale_x_reverse(breaks = sort(unique(results_shells$max_radius))) +
  coord_cartesian(ylim = c(0, max(results_shells$estimated_k) + 1)) +
  labs(
    title = "Estimated number of clusters vs max radius",
    x = "Maximum radius",
    y = "Estimated # of clusters",
    color = ""
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )
```

As the maximum radius decreases and the shells get closer together, it becomes more difficult to predict the accurate number of clusters. This is why we see better predictions with larger radius values. The algorithm’s failure point happens when the shells are so close together that the points may overlap, possibly leading to either merged clusters. This result would change if I used a different d_threshold, for example 0.8 or 1.2. Lower thresholds like 0.8 would restrict the graph such that only closer points become connected (fewer points are connected in general compared to d_threshold =1). Therefore, lower threshold values may overestimate the number of clusters when clusters are split into smaller sections based on less connectivity. In contrast, higher threshold values like 1.2 may increase connectivity in the graph, allowing for clusters to potentially merge and result in lower cluster estimates. A lower threshold may help distinguish clusters when the radius is small because it restricts connectivity to nodes in the graph that have a stronger connection, compared to a larger threshold that may be more warranted with larger radii.

However, as the shell radius gets larger the distance between points in the shell naturally decreases - same number of points distributed over a larger volume/surface area. Therefore, spectral clustering becomes challenging and the effect described above is less obvious, based on the hw assignment itself.

